---
title: "Analysing GDP and Income Across Countries"
author: "Lakshmi Bobbillapati, Stanford Finney, Laura Ludwig, Ryan Talk"
date: "October 17, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is where we will write an introduction to the problem and analysis we are completing.

## Data cleaning

In this section, we will need to present code that we use to gather and merge the two data sets (so Professor Woo can see it). We also need to include several text sentences before or after the code chunk to explain what is being done. 

Reminder, code chunks are like this:
```{r cars}
summary(cars)
```

## Addressing Questions of Interest 

We will need to provide code addressing each question that is outlined in the assignment: 

1. Merge the data based on the country shortcode. How many of the IDs match? 
Insert code chunk, followed by answer here.

2. Sort the data frame in ascending order by GDP (so United States is last). What is the 13th country in the resulting data frame?
Insert code chunk, followed by answer here.

3. What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups? 
Insert code chunk, followed by answer here.

4. Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.
Insert code chunk, followed by answer here.

5. Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP?
Insert code chunk, followed by answer here.

## Conclusion
Here we will need to tie together the results from our analysis.


