---
title: "Analysing GDP and Income Across Countries"
author: "Lakshmi Bobbillapati, Stanford Finney, Laura Ludwig, Ryan Talk"
date: "October 17, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
load("merged.data")
load("gdp.data")
load("edstats.data")
```

## Introduction

This is where we will write an introduction to the problem and analysis we are completing.

## Data cleaning

In this section, we will need to present code that we use to gather and merge the two data sets (so Professor Woo can see it). We also need to include several text sentences before or after the code chunk to explain what is being done. 

Reminder, code chunks are like this:

Clean GDP
```{r cleanGDP}
library(readr)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

gdp <- read.csv("GDP.csv", skip = 3)
# Remove extraneous columns
gdp[c(3,6:10)] <- NULL
names(gdp)[c(1, 4)] <- c("CountryCode", "US.dollars")
# Remove Blank Rows
gdp <- gdp[-which(gdp$CountryCode==""),]

gdp$US.dollars <- gsub(",", "", gdp$US.dollars)
  ## ".." in the US Dollars column represents not available. as.Numeric automatically converts these to NA
gdp$US.dollars <- as.numeric(as.character(gdp$US.dollars))
gdp$Ranking <- as.numeric(as.character(gdp$Ranking))

# Data for the Regions
gdp.stats <- gdp[which(is.na(gdp$Ranking) & !is.na(gdp$US.dollars)),]
# Data For each country
gdp <- gdp[-which(is.na(gdp$Ranking) & !is.na(gdp$US.dollars)),]

save(gdp, gdp.stats, file = "gdp.data")
```

Clean EDSTATS

```{r cleanEDSTATS}
edstats <- read.csv("EDSTATS.csv", stringsAsFactors = FALSE)
# Remove extraneous columns
edstats <- edstats[1:3]

# Remove blank rows
edstats <- edstats[-which(edstats$Income.Group==""),]
# Simplify group names
edstats$Income.Group <- revalue(edstats$Income.Group, c("Low income" = "LI", 
                                                        "Lower middle income" = "LMI", 
                                                        "High income: nonOECD" = "HI_nonOECD", 
                                                        "Upper middle income" = "UMI", 
                                                        "High income: OECD" = "HI_OECD"))

save(edstats, file = "edstats.data")
```

## Addressing Questions of Interest 

``` {r numNA}
#Number of NA's for GDP:
sum(is.na(gdp$US.dollars))
sum(is.na(gdp$Ranking))
```
We will need to provide code addressing each question that is outlined in the assignment: 

1. Merge the data based on the country shortcode. How many of the IDs match? 
Insert code chunk, followed by answer here.
```{r question1}
# Merge the data
merged <- merge(gdp, edstats, by = "CountryCode")

# Number of common Country Codes
length(intersect(gdp$CountryCode, edstats$CountryCode))
```


2. Sort the data frame in ascending order by GDP (so United States is last). What is the 13th country in the resulting data frame?
Insert code chunk, followed by answer here.
``` {r question2}
# Sort Merged data by GDP and retrieve 13th row
merged.sortGDP <- merged[order(merged$US.dollars),]
merged.sortGDP[13,]
```

3. What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups? 
Insert code chunk, followed by answer here.
``` {r question3}
# Get the Average GDP for High Income OECD
mean(merged$US.dollars[which(merged$Income.Group=="HI_OECD")])

# Get the Average GDP for High Income nonOECD
mean(merged$US.dollars[which(merged$Income.Group=="HI_nonOECD")], na.rm = TRUE)
```

4. Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.
Insert code chunk, followed by answer here.

``` {r question4}
library(ggplot2)
# Plot GDP for all the countries, color-coded by income group
ggplot(merged, aes(x =CountryCode, y = US.dollars, color=Income.Group)) +geom_point()
```

5. Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP?
Insert code chunk, followed by answer here.

```{r question5}
# Cut the GDP Ranking into 5 separate quantile groups
quantiles <- list()
for(i in 1:5){
  quantiles[[i]] <- merged[((i-1)*38+1):(i*38),]
}

# Make a table versus Income.Group.
ccIncomeGroupTable <- with(quantiles[[1]], table(CountryCode, Income.Group))

# How many countries are in the Lower Middle Income, but in the top 38 nations with highest GDP?
sum(quantiles[[5]]$Income.Group=="LMI")
```

## Conclusion
Here we will need to tie together the results from our analysis.


