---
title: "Analysing GDP and Income Across Countries"
author: "Lakshmi Bobbillapati, Stanford Finney, Laura Ludwig, Ryan Talk"
date: "October 17, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Introduction

The World Bank's Open Data initiative is inteded to provide all users with access to World Bank Data.  
Gross Domestic Product (GDP) is the single most commonly referenced figure to cover the entirety of a national economy 
and the trajectory it is on in a single statistic.

The World Bank Education Statistics is the comprehensive data and analysis source for key topics in education such as Income group, region, latest population census.
The Organization for Economic Cooperation and Development (OECD) is a unique forum where the governments of 34 democracies with market economies work with each other to promote economic growth, prosperity and development.  
The income groups are categorized based on High income with OECD or Non-OECD, lower / upper middle income and low income.


The goal of this case study is to understand the relationship between GDP and the education income groups.  
Does countries with high GDP have low income groups (with / without OECD) or vice versa.

## Data cleaning

In order to complete our analysis, several data sets are needed. These are gathered from two locations online, and then need to be cleaned and merged in order to be useable in the analysis. 

The first data set relates to GDP data for many countries. The key in this data is a three-character country code, and the resulting output data includes this code, the GDP in US dollars, and the GDP ranking. This data set also included some summary measurements about regional statistics, which were excluded for the purpose of this analysis.

```{r cleanGDP}
library(readr)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

# Read in and store data from csv file
gdp <- read.csv("GDP.csv", skip = 3)
# Remove extraneous columns
gdp[c(3,6:10)] <- NULL
names(gdp)[c(1, 4)] <- c("CountryCode", "US.dollars")
# Remove Blank Rows
gdp <- gdp[-which(gdp$CountryCode==""),]

# Remove the comma formatting so a double is stored 
gdp$US.dollars <- gsub(",", "", gdp$US.dollars)

# ".." in the US Dollars column represents not available. as.Numeric automatically converts these to NA
gdp$US.dollars <- as.numeric(as.character(gdp$US.dollars))
gdp$Ranking <- as.numeric(as.character(gdp$Ranking))

# Segregate data for the Regions
gdp.stats <- gdp[which(is.na(gdp$Ranking) & !is.na(gdp$US.dollars)),]
# Segregate data For each country
gdp <- gdp[-which(is.na(gdp$Ranking) & !is.na(gdp$US.dollars)),]

#save a file that includes the ouput of the above code
save(gdp, gdp.stats, file = "gdp.data")
```

The second data set relates to income to many countries. The key in this data is the same three-character country code as the first data set. The resulting output data includes this code, the country name and the income group. The Income group was codified as it was brought in, and the data set was saved for further use.


```{r cleanEDSTATS}
edstats <- read.csv("EDSTATS.csv", stringsAsFactors = FALSE)
# Remove extraneous columns
edstats <- edstats[1:3]

# Remove blank rows
edstats <- edstats[-which(edstats$Income.Group==""),]
# Simplify group names
edstats$Income.Group <- revalue(edstats$Income.Group, c("Low income" = "LI", 
                                                        "Lower middle income" = "LMI", 
                                                        "High income: nonOECD" = "HI_nonOECD", 
                                                        "Upper middle income" = "UMI", 
                                                        "High income: OECD" = "HI_OECD"))

save(edstats, file = "edstats.data")
```

## Addressing Questions of Interest 

``` {r numNA}
#Number of NA's for GDP:
sum(is.na(gdp$US.dollars))
sum(is.na(gdp$Ranking))
```
 

1. Merge the data based on the country shortcode. How many of the IDs match? 

```{r question1}
# Merge the data
merged <- merge(gdp, edstats, by = "CountryCode")
```


The number of common country codes (matching IDs) is 
```{r q1answer} 
length(intersect(gdp$CountryCode, edstats$CountryCode))
```


2. Sort the data frame in ascending order by GDP (so United States is last). What is the 13th country in the resulting data frame?

``` {r question2}
# Sort Merged data by GDP and retrieve 13th row
merged.sortGDP <- merged[order(merged$US.dollars),]
```
The country with the 13th smallest GDP is 
```{r q2answer} 
merged.sortGDP[13,]
```

3. What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups? 

The average GDP ranking for "High income: OECD" is
``` {r q3answer1}
# Get the Average GDP for High Income OECD
mean(merged$US.dollars[which(merged$Income.Group=="HI_OECD")])
```
The average GDP ranking for "High income: nonOECD" is
``` {r q3answer2}
# Get the Average GDP for High Income nonOECD
mean(merged$US.dollars[which(merged$Income.Group=="HI_nonOECD")], na.rm = TRUE)
```

4. Plot the GDP for all of the countries. 
Below is a plot showing the GDP for all countries, with colors coded to the Income Group.
``` {r question4}
library(ggplot2)
# Plot GDP for all the countries, color-coded by income group
ggplot(merged, aes(x =CountryCode, y = US.dollars, color=Income.Group)) +geom_point()
```

5. Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP?

```{r question5}
# Cut the GDP Ranking into 5 separate quantile groups
quantiles <- list()
for(i in 1:5){
  quantiles[[i]] <- merged[((i-1)*38+1):(i*38),]
}

# Make a table versus Income.Group.
ccIncomeGroupTable <- with(quantiles[[1]], table(CountryCode, Income.Group))
```
```{r q5answer}
# How many countries are in the Lower Middle Income, but in the top 38 nations with highest GDP?
sum(quantiles[[5]]$Income.Group=="LMI")
```
are in the Lower Middle Income, but in the top 38 nations with highest GDP

## Conclusion
With the data gathered by the World Bank’s Open Data Initiative we can view correlations between countries’ GDP and educational statistics, by parsing, cleaning, and creating visual aids pertaining countries’ GDP and educational statistics. Utilizing the data gathered by the World Bank, we are able to clean and sort data, as well as extrapolate information to address questions of interest. 

For the GDP data set we take the three-character country code, GDP in dollars, and the GDP ranking, and for the educational data set we use he three-character country code, country’s name and income group. Utilizing the two data sets, and the R packages tools we are able to create graphs, and view correlations between countries, GDP, and education.

